---
title: "CITES trade networks in DiagrammeR"
author: "Tamora James"
date: "07/11/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(DiagrammeR)
library(DiagrammeRsvg) # required for export_graph()
library(RColorBrewer)
pal <- brewer.pal(n = 8, "Dark2")
```

The trade statistics used in this demonstration are derived from the CITES Trade Database[^1]. Under CITES rules, trade in CITES listed species must be recorded by exporting and importing countries. I downloaded a comparative tabulation for recorded trade in live _Amazona barbadensis_ parrots between 1975 and 2017.

```{r data-import}
citesData <- read.csv("data/comptab_Amazona_barbadensis_1975-2017.csv",
                      stringsAsFactors = FALSE)
desc <- "Trade in Amazona barbadensis 1975-2017"
```

The first task is to set up nodes for trading countries:

```{r node-setup}
## Get list of unique node names for all trade roles
node_names <- citesData %>%
  select(Importer, Exporter, Origin) %>%
  unlist(use.names = FALSE) %>%
  unique()
node_names <- node_names[node_names != ""]

## Create node data frame for trade nodes
nodes <- create_node_df(
    n = length(node_names),
    type = node_names,
    label = node_names,
    ## Additional attributes: trade roles
    exporter = node_names %in% citesData$Exporter,
    importer = node_names %in% citesData$Importer,
    origin = node_names %in% citesData$Origin
)
head(nodes)
```

Next, I created edge relationships between trading countries, differentiating between reported exports and imports:

```{r edge-setup}
## Create edges for each trade relationship
edges_e <-
    create_edge_df(
        from = nodes$id[match(citesData$Exporter, nodes$type)],
        to = nodes$id[match(citesData$Importer, nodes$type)],
        rel = "export",
        data = citesData$Exporter.reported.quantity,
        source = citesData$Source,
        penwidth = ifelse(is.na(citesData$Exporter.reported.quantity), 1,
                          log(citesData$Exporter.reported.quantity+1)),
        color = pal[1]
    )

edges_i <-
    create_edge_df(
        from = nodes$id[match(citesData$Exporter, nodes$type)],
        to = nodes$id[match(citesData$Importer, nodes$type)],
        rel = "import",
        data = citesData$Importer.reported.quantity,
        source = citesData$Source,
        penwidth = ifelse(is.na(citesData$Importer.reported.quantity), 1,
                          log(citesData$Importer.reported.quantity+1)),
        color = pal[2]
    )

edges <- combine_edfs(edges_e,
                      edges_i)

head(edges)
```

Now I can combine the nodes and edges to create and render a graph:

```{r create-graph-1}
cites_graph <- create_graph(nodes_df = nodes,
                            edges_df = edges
                            )

render_graph(cites_graph,
             title = desc,
             layout = "neato"
             )
```

But this produces a crazy graph because there are many edges between node pairs.

```{r, include = FALSE}
## Filter to only reported imports
cites_graph %>%
    select_edges(conditions = rel == "import") %>%
    render_graph()
```

Clearly a different approach is needed to make the diagram more manageable.
I'll combine the edges for trade between the same countries, using the maximum of the reported quantities for exports and imports.

```{r}
## Create export summary data
exportSummary <- citesData %>%
  group_by(Exporter, Importer) %>%
  summarise(quantity = sum(max(Importer.reported.quantity, Exporter.reported.quantity, na.rm = TRUE)))

## Create edges for trade summary
edges_e2 <-
  create_edge_df(
    from = nodes$id[match(exportSummary$Exporter, nodes$type)],
    to = nodes$id[match(exportSummary$Importer, nodes$type)],
    rel = "export",
    quantity = exportSummary$quantity,
    penwidth = log(exportSummary$quantity+1),
    color = pal[1]
    )
```

(It would be fun to calculate the distance between countries and set the edge lengths accordingly but for the moment I'll allow the algorithm to select default values.)

```{r}
cites_graph2 <- create_graph(nodes_df = nodes,
                             edges_df = edges_e2
)

render_graph(cites_graph2,
             title = desc,
             layout = "nicely"  # options are: nicely, circle, tree, kk, and fr
)

export_graph(cites_graph2, file_type = "png", file_name = "cites_graph.png")

```

[^1]: [CITES Trade Database, UNEP World Conservation Monitoring Centre, Cambridge, UK](https://trade.cites.org/en/cites_trade/).
